Decompositions of relative Chevalley groups

### 2015-10-07 dennis-vaserstein.tex
Improved 4.2
Moved Gauss and DV theorems to the introduction

The main result of this subsection is the following theorem which is a relative version of Dennis---Vaserstein theorem (cf. \cite[Lemma~2.1]{ST76} and \cite[Theorem~2.5]{St78}).
Notice that M.~Stein uses the first of the above $4$ equivalent ways of writing down the decomposition while A.~Suslin and M.~Tulenbaev use the 4th.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi$ be a reduced irreducible root system not equal to $\rE_8$ and let $\{ r, s \}$ denote the labels of some pair of distinct endnodes of the Dynkin diagram of $\Phi$
such that $\{ r, s\} \neq \{ 1,2 \}$ for $\Phi=\rE_\ell$. Denote by $d$ the distance between $r$ and $s$ on the Dynkin diagram.
Then the relative elementary subgroup $\E(\Phi, R, I)$ admits Dennis---Vaserstein decomposition, i.\,e. $\E(\Phi, R, I) = A_{rs}$ 
under the following assumption on $I$:
\begin{enumerate}
 \item $\asr(I) \leq d$ for $\Phi=\rF_4$ or $\Phi=\rE_\ell$, $\{r, s\} = \{1, \ell \}$,
 \item $\sr(I) \leq d$, otherwise.
\end{enumerate}

%The relative elementary subgroup $\E(\Phi, R, I)$ admits Dennis---Vaserstein decomposition, i.\,e. $\E(\Phi, R, I) = A_{rs}$ under the following assumptions on $I$ and $\Phi$.
%\[\begin{array}{l@{\qquad}c@{\quad}c@{\qquad}l}
%\Phi                                 & s    & r      & \text{ring condition} \\ \hline\vphantom{\Bigl(}
%\rA_\ell,\ \ell\geqslant 2           & 1    & \ell   & \sr(I) \leqslant \ell-1 \\    
%\rB_\ell,\ \ell\geqslant 2           & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
%\rC_\ell,\ \ell \geqslant 2          & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
%\rD_\ell,\ \ell\geqslant 4           & 1    & \ell   & \sr(I) \leqslant \ell-2 \\ 
%\rD_\ell,\ \ell\geqslant 4           & \ell & \ell-1 & \sr(I) \leqslant 2      \\
%\rE_\ell,\ \ell=6,7                  & \ell & 2      & \sr(I) \leqslant \ell-3 \\ 
%\rE_\ell,\ \ell=6,7                  & \ell & 1      & \asr(I)\leqslant \ell-2 \\ 
%\rF_4                                & 4    & 1      & \asr(I)\leqslant 3 \end{array}\]
\end{thm}
Using identities~\ref{rel:h-comm}--\ref{rel:h-w} it is easy to show that Dennis---Vaserstein decomposition is a consequence of Gauss decomposition (see \cref{thm:srRI1}).
In particular, for rings of stable rank $1$ and groups of rank $2$ (e.\,g. $\Phi=\rG_2$) it is preferable to use Gauss decomposition since it is more accurate.

>>>>

%The relative elementary subgroup $\E(\Phi, R, I)$ admits Dennis---Vaserstein decomposition, i.\,e. $\E(\Phi, R, I) = A_{rs}$ under the following assumptions on $I$ and $\Phi$.
%\[\begin{array}{l@{\qquad}c@{\quad}c@{\qquad}l}
%\Phi                                 & s    & r      & \text{ring condition} \\ \hline\vphantom{\Bigl(}
%\rA_\ell,\ \ell\geqslant 2           & 1    & \ell   & \sr(I) \leqslant \ell-1 \\    
%\rB_\ell,\ \ell\geqslant 2           & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
%\rC_\ell,\ \ell \geqslant 2          & 1    & \ell   & \sr(I) \leqslant \ell-1 \\
%\rD_\ell,\ \ell\geqslant 4           & 1    & \ell   & \sr(I) \leqslant \ell-2 \\ 
%\rD_\ell,\ \ell\geqslant 4           & \ell & \ell-1 & \sr(I) \leqslant 2      \\
%\rE_\ell,\ \ell=6,7                  & \ell & 2      & \sr(I) \leqslant \ell-3 \\ 
%\rE_\ell,\ \ell=6,7                  & \ell & 1      & \asr(I)\leqslant \ell-2 \\ 
%\rF_4                                & 4    & 1      & \asr(I)\leqslant 3 \end{array}\]

Using identities~\ref{rel:h-comm}--\ref{rel:h-w} it is easy to show that Dennis---Vaserstein decomposition is a consequence of Gauss decomposition (see \cref{thm:srRI1}).
In particular, for rings of stable rank $1$ and groups of rank $2$ (e.\,g. $\Phi=\rG_2$) it is preferable to use Gauss decomposition since it is more accurate.

### 2015-10-07 gauss.tex
Added reference on Abe '69
Fixed references to Abe
Fixes from 05.10.15 meeting
Moved Gauss and DV theorems to the introduction

The proof of Gauss decomposition presented below is similar to the absolute case (cf.~\cite[Theorem~5.1]{Sm12}).

\begin{prop}\label{thm:Gauss}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ and let $\Delta_1, \Delta_\ell$ be
two reductive subsystems of $\Phi$ corresponding to the endnodes of the Dynkin diagram of $\Phi$.
Suppose that both relative elementary subgroups $\E(\Delta_i, R, I)$, $i=1,\ell$ admit triangular decomposition with $N$ triangular factors:
\[ \E(\Delta_i, R, I) = \Hh(\Delta_i, R, I) \cdot \U(\Delta^+_i, I) \cdot \U(\Delta^-_i, I) \cdot \ldots \cdot \U(\Delta^\pm_i, I),\quad i=1,\ell. \]
Then the group $\E(\Phi, R, I)$ admits the decomposition with the same number of factors:
\[ \E(\Phi, R, I) = \Hh(\Phi,R,I) \cdot \U(\Phi^+,I) \cdot \U(\Phi^-, I) \cdot \ldots \cdot \U(\Phi^\pm,I). \]
\end{prop}
\begin{proof}
Denote by $Y$ the product of subgroups in the right-hand side of the above equality.
To show that $Y=\E(\Phi,R,I)$ it suffices to check that
\begin{enumerate}
\item $Y$ is normalized by $\E(\Phi,R)$, i.\,e. $Y^{\E(\Phi,R)}\subseteq Y$;
\item there exists a set $X$ generating $\E(\Phi,R,I)$ as a \emph{normal} subgroup of $\E(\Phi,R)$ such that $XY\subseteq Y$.
\end{enumerate}
To prove the first assertion it suffices to show that $Y^{x_\alpha(\xi)} \subseteq Y$ for any $\alpha\in \pm \Pi$, $\xi\in R$.
Fix a root $\alpha\in\pm\Pi$. Clearly, $\alpha \in \Delta_i$ for $i=1$ or $i=\ell$ and we can expand $Y$ as follows:
\[ Y=\Hh(\Phi,R,I) \cdot \U(\Delta_i^+,I) \cdot \ldots \cdot \U^\pm(\Delta_i,I) \cdot \U(\Sigma_i^+,I) \cdot \ldots \cdot \U(\Sigma_i^\pm,I). \]
For every  $h \in \Hh(\Phi, R, I)$ one has for some $s\in I$
\[ x_\alpha(\xi)\cdot h = h\cdot x_\alpha((1+s)\xi)=h\cdot x_\alpha(\xi)\,x_\alpha(s\xi). \]

Therefore by the assumption of the theorem
\begin{multline*}
Y^{x_\alpha(\xi)} \subseteq \Hh(\Phi,R,I) \cdot x_\alpha(-\xi) X_{\alpha}(I) \cdot \E(\Delta_i,R,I)\cdot x_\alpha(\xi) \cdot \U(\Sigma^+_i,I) \cdot \ldots \cdot \U(\Sigma^\pm_i,I) =\\
= \Hh(\Phi, R, I) \cdot \E(\Delta_i,R,I) \cdot \U(\Sigma^+_i,I) \cdot \ldots \cdot \U(\Sigma^\pm_i,I) = Y.
\end{multline*}

@@

\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley group admits Gauss decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \cdot \U(\Phi,I) \cdot \U(\Phi^-,I) \cdot \U(\Phi,I). \]
\end{thm}
\begin{proof}
In view of the above proposition it suffices to prove that Gauss decomposition of length $3$ holds for $\Phi=\rA_1$.

>>>>

The proof of Gauss decomposition presented below is similar to the absolute case (cf. with the proof of~\cite[Theorem~5.1]{Sm12}).
\begin{prop}\label{thm:Gauss}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ and let $\Delta_1, \Delta_\ell$ be
two reductive subsystems of $\Phi$ corresponding to the endnodes of the Dynkin diagram of $\Phi$.
Suppose that both relative elementary subgroups $\E(\Delta_i, R, I)$, $i=1,\ell$ admit triangular decomposition with $N$ triangular factors:
\[ \E(\Delta_i, R, I) = \Hh(\Delta_i, R, I) \cdot \U(\Delta^+_i, I) \cdot \U(\Delta^-_i, I) \cdot \ldots \cdot \U(\Delta^\pm_i, I),\quad i=1,\ell. \]
Then the group $\E(\Phi, R, I)$ admits the decomposition with the same number of factors:
\[ \E(\Phi, R, I) = \Hh(\Phi,R,I) \cdot \U(\Phi^+,I) \cdot \U(\Phi^-, I) \cdot \ldots \cdot \U(\Phi^\pm,I). \]
\end{prop}
\begin{proof}
Denote by $Y$ the product of subgroups in the right-hand side of the above equality.
To show that $Y=\E(\Phi,R,I)$ it suffices to check that
\begin{enumerate}
\item $Y$ is normalized by $\E(\Phi,R)$, i.\,e. $Y^{\E(\Phi,R)}\subseteq Y$;
\item there exists a set $X$ generating $\E(\Phi,R,I)$ as a \emph{normal} subgroup of $\E(\Phi,R)$ such that $XY\subseteq Y$.
\end{enumerate}
To prove the first assertion it suffices to show that $Y^{x_\alpha(\xi)} \subseteq Y$ for any $\alpha\in \pm \Pi$, $\xi\in R$.
Fix a root $\alpha\in\pm\Pi$. Clearly, $\alpha \in \Delta_i$ for $i=1$ or $i=\ell$ and we can expand $Y$ as follows:
\[ Y=\Hh(\Phi,R,I) \cdot \U(\Delta_i^+,I) \cdot \ldots \cdot \U^\pm(\Delta_i,I) \cdot \U(\Sigma_i^+,I) \cdot \ldots \cdot \U(\Sigma_i^\pm,I). \]
For every  $h \in \Hh(\Phi, R, I)$ one has for some $s\in I$
\[ x_\alpha(\xi)\cdot h = h\cdot x_\alpha((1+s)\xi)=h\cdot x_\alpha(\xi)\,x_\alpha(s\xi). \]

Therefore by the assumption of the theorem
\begin{multline*}
Y^{x_\alpha(\xi)} \subseteq \Hh(\Phi,R,I) \cdot x_\alpha(-\xi) \cdot X_{\alpha}(I) \cdot \E(\Delta_i,R,I)\cdot x_\alpha(\xi) \cdot \U(\Sigma^+_i,I) \cdot \ldots \cdot \U(\Sigma^\pm_i,I) =\\
= \Hh(\Phi, R, I) \cdot \E(\Delta_i,R,I) \cdot \U(\Sigma^+_i,I) \cdot \ldots \cdot \U(\Sigma^\pm_i,I) = Y.
\end{multline*}

@@

\begin{proof}[Proof of \cref{thm:srRI1}]
In view of the above proposition it suffices to prove that Gauss decomposition of length $3$ holds for $\Phi=\rA_1$.

### 2015-10-07 paper.tex
Moved Gauss and DV theorems to the introduction

\begin{abstract} 
We formulate and prove relative versions of Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
As an application we obtain upper estimates for the width of principal congruence subgroups in terms of several families of generators.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots.
Denote by $G=\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over $R$ and by $\E(\Phi, R)$ its \emph{elementary subgroup}, 
i.\,e. the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (also called \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

The main goal of the present paper is to study parabolic factorizations of Chevalley groups and their analogues for principal congruence subgroups $\G(\Phi, R, I)$.
By a parabolic factorization we mean a decomposition of $G$ into a product of several parabolic and unipotent subgroups.
To be more specific, we study relative analogues of the following three kinds of decompositions:
\begin{itemize}
\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

@@

It is also classically known that over an arbitrary field $F$ the group $\SL_{n+1}(F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, from the classical Dennis---Vaserstein theorem (see~\cite{Va69} and \cite[Lemma~2.1]{ST76})
it follows that $\SL_{n+1}(R, I)$ is a product of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.
Using another Dennis---Vaserstein type factorization we obtain the following result.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$ then the group $\Epin_{2\ell}(R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ copies of its subgroups of type $\rA_{\ell-1}$.
\end{thm}

>>>>

\begin{abstract} 
We formulate and prove relative versions of several classical decompositions known in the theory of Chevalley groups over commutative rings. % Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
As an application we obtain upper estimates for the width of principal congruence subgroups in terms of several families of generators.
\end{abstract}

@@

\section{Introduction}\label{sec:intro}
%Let $R$ be a commutative unital ring and $\Phi$ be a reduced irreducible root system of rank $\ell$ with a fixed basis of simple roots.
Let $\G(\Phi, R)$ the split simply-connected Chevalley group of type $\Phi$ over arbitarary commutative ring $R$ and let $\E(\Phi, R)$ be its \emph{elementary subgroup} (see e.\,g.~\cite{St78, VP, S}).
%i.\,e. the subgroup generated by root unipotents $x_\alpha(\xi)$, $\alpha\in\Phi$, $\xi\in R$ see~\cite{VP, St78, S}.
For an ideal $I \trianglelefteq R$ we denote by $\G(\Phi, R, I)$ the \emph{relative Chevalley group} (aka \emph{principal congruence subgroup}) of level $I$, 
i.\,e. the kernel of the map $\G(\Phi, R)\rightarrow\G(\Phi, R/I)$ induced by the canonical projection $R\rightarrow R/I$.

The main goal of the present paper is to obtain relative analogues of several classical parabolic factorizations of Chevalley groups (i.\,e. factorizations into a product of several parabolic and unipotent subgroups)
%To be more specific, we study relative analogues of the following three kinds of decompositions:
%\begin{itemize}
%\item \emph{Gauss decomposition,} i.\,e. a decomposition of the form $\E(\Phi,R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup,
%      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
%\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
%      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
%\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$,
%      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
%\end{itemize}
%TODO: Insert some links to VIP's papers :)

@@

Our first result is a relative version Gauss decomposition.
All the notation for the subgroups appearing in this theorem and the other theorems stated below is explained in~\ref{sec:prelim}.
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley $\E(\Phi, R, I)$ group admits the following decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \cdot \U(\Phi,I) \cdot \U(\Phi^-,I) \cdot \U(\Phi,I). \]
\end{thm}
The special case $I=R$ of the \cref{thm:srRI1} was obtained by the second named author in~\cite{Sm12}.
We also note that in special case of a semilocal ring $R$ the above theorem was proved by E.~Abe and K.~Suzuki, see~\cite[Proposition~2.3]{AbeSuzNormalSubgroups}.

Our next main result is a relative Dennis---Vaserstein decomposition which holds for a broader class of groups of rings as compared to \cref{thm:srRI1}.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$ the distance between which equals $d$.
Assume that $\Phi$ and $I$ satisfy either of the following assumptions:
\begin{enumerate}
 \item $\asr(I) \leq d$ for $\Phi=\rF_4$ or $\Phi=\rE_\ell$, $\{r, s\} = \{1, \ell \}$,
 \item $\sr(I) \leq d$, otherwise.
\end{enumerate}
Then $\E(\Phi, R, I)$ factors as $\EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s) \cdot \EP_s(R, I).$ \end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to M.~Stein's Theorems~2.5, 4.1, see~\cite{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).

The condition $\asr(I)$ used in the statement of \cref{thm:DennisVaserstein} is a relative version of the \emph{absolute stable range} of D.~Estes and J.~Ohm which does not depend on the ambient ring $R$ (see section~\ref{sec:rel-asr}).

@@

It is also classically known that over an arbitrary field $F$ the group $\SL_{n+1}(F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, from the classical Dennis---Vaserstein theorem (see~\cite{Va69} and \cite[Lemma~2.1]{ST76})
it follows that $\SL_{n+1}(R, I)$ is a product of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.
Using another Dennis---Vaserstein type factorization we obtain the following result.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$ then the group $\Epin_{2\ell}(R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ copies of its (regularly embedded) subgroups of type $\rA_{\ell-1}$.
\end{thm}
